Human Factors Influencing Visual Statistical Inference

### 2013-07-19 sociology_paper.tex
new figure on education effect added
still working on section 2

There are a variety of ways that data can be plotted and enhanced. The choice of plot that is used as the visual test statistic should be closely associated with the problem being investigated. For example, side-by-side boxplots may be more appropriate than side-by-side dotplots, when one variable is categorical. Enhancements, such as a line overlaying a scatterplot to represent a fitted model, might be appropriate for studying some types of problems.  Some choices of visual test statistics provide better results than others when studying the same problem, which is discussed in \cite{heike:2012}.  This makes sense, because some plot types will make it easier for see some structure. It also presents as a flip-side to visual inference, that the same methods could be used to determine the optimal plot choice.

Table \ref{tbl:visual_stat} shows the visual test statistics used in the Turk studies completed so far. Side-by-side boxplots were used in experiment 1 to study the importance of a categorical variable in a regression model. For both experiments 2 and 3 a scatterplot is used but a regression line fitted through the points is overlaid for experiment 2. It is easier to spot the slope of the line compared to just the scatterplot itself. But people are better in noticing the unusual pattern in the data which is some contamination purposefully added for experiment 3. Experiment 4 compared polar coordinates to euclidean coordinates on the reading of structure in a barchart.  In addition, this experiment also examined the style of plot for a bivariate data problem: a scatterplot, side-by-side dot plots, overlaid densities or side-by-side boxplots. .... expand description of experiments here...

@@

The visual test statistic is chosen so that it displays a specific pattern in case the null hypothesis is not true. Thus the most important factor that help an observer to correctly evaluate a lineup is the presence of any detectable signal in the data. On the other hand, if the null hypothesis is not true, visual test statistic should not display any distinguishable pattern. In fact, some of the null plots may appear to be the most different plot in a lineup influencing the observer to chose a plot different than actual data plot. This is an elegant feature of the lineup. It force the observer to chose a wrong plot when the null hypothesis should not be rejected.

@@

\subsection{Question that Human Observer Answers} The researcher knows about the underlying hypothesis but the observer does not necessarily know the underlying details of the lineup. So, the researcher needs to ask a question to the observer to answer while evaluating the lineup. This question should provide the observer a little clue so that the answer reflects the hypothesized patterns in the actual plot. For example Table \ref{tbl:visual_stat} shows the questions asked for the simulation experiments considered for this study. Notice that for case 1 if the observer can identify the actual plot in the lineup that should indicate that the plot chosen has the most vertical difference between groups A and B which is exactly what the researchers intend to examine. Similarly for case 2, a correct evaluation would indicate that the slope is different than the slope that may show up just from randomness.

@@

\subsection{Demographics of the Observer}  Some of the demographics of the observer such as age, gender, education level and geographical location may have effect on how an observer would examine the lineup. This may produce some variability in the performance of the observer. For example, a high school student may not respond same as a well educated person while evaluating a lineup. There may be variations in different age groups as well. Investigation on subjects with a well variety of age groups is necessary to study this variability.

@@

\subsection{Learning Trend of the Observer} When an observer evaluates a lineup he or she may learn something from the experiences of the earlier observations. If the same observer is shown another lineup the learning from previous evaluation may help. The more evaluations an observer makes the more skillful the observer may become. This learning trend may or may not be significant overall but it may influence the performance of the observer in some way.

@@

\subsection{Location of Actual Plot in the Lineup} The actual data plot is placed in a random spot in a lineup. While evaluating the lineup some people may start looking from some specific part of the lineup. With the help of eye-tracking equipments \citet{zhao:2012} tracked the observers' eyes to see how they went through the plots in a lineup to come to their answers. The results suggest that people have particular methods of reading lineups. Some people read lineups from left to right direction while some read from upward to downward. Some people start looking from the center of the lineup while others start from the top left corner. In the earlier phase of the exercise, the observer tend to scan the plots and start comparing plots to make a final decision. Beside right-left or up-down directions observers show some diagonal movement too.

@@

\subsection{Individual Skill or Ability of the Observer} Each person is different from others in some way. For example, in a controlled experimental study \citet{zhao:2012}  noticed that some people spent a lot of time to decide no matter whether the lineup is difficult or easy while some simply glanced at lineups to make a decision. This influences the response of the observer. Also subject specific variation in the power of visual test is observed in \cite{majumder:2013}.

@@

\begin{figure}[htbp] 
   \centering
%   \includegraphics[width=6.5in]{learning_trend.pdf} 
    \includegraphics[width=3.5in]{practical_impact_graduate.pdf} 
   \caption{Proportion of correct responses due to graduate degree as compared to high school degree for an 18-25 year old female in the United States. Even though graduate degree is statistically significant, the largest difference in proportion correct is  0.045 which is very negligible. The difference diminishes as we move away one or two standard deviations ($\sigma_{\ell} =2.293$) of lineup variability.}
   \label{fig:practical_impact_graduate}
\end{figure}

>>>>

There are a variety of ways that data can be plotted and for plots to be enhanced by axes, legends or representations of models. The choice of plot chosen as the visual test statistic, primarily should be appropriate for the problem being investigated. For example, for studying the association between two quantitative variables a scatterplot is usually optimal. Enhancements, such as a line overlaying a scatterplot to represent a fitted model, might be appropriate for studying models for the data.  Some choices of visual test statistics provide better results than others when studying the same problem, which is discussed in \cite{heike:2012}.  This makes sense, because some plot types will make it easier for certain structures to be seen.

Table \ref{tbl:visual_stat} shows the large variety of visual test statistics used in the Turk studies completed so far. These experiments were conducted to examine many different visual tasks. Side-by-side boxplots were used in experiment 1 to study the importance of a categorical variable in a regression model. For both experiments 2 and 3, a scatterplot is the visual test statistic, but a regression line fitted through the points is overlaid for experiment 2. The experiments studied slightly different problems, for which the difference in plot type was important. Experiment 4 compared polar coordinates to euclidean coordinates on the reading of structure in a barchart.  Experiment 5 examined the style of plot for a bivariate data problem, by comparing results when using a scatterplot, side-by-side dot plots, overlaid densities or side-by-side boxplots. Experiment 6 studied the difference between variations of boxplots. Experiment 7 studied the effects of high dimension low sample size on the separation of groups in low-dimensional projections. The sine illusion, where people read the strength of differences between curves as different when there is a strong sinusoidal structure, is examined in experiment 8. Experiment 9 was conducted to test if there was any structure in an RNA-seq data set measured on soybeans, and used interaction plots classically constructed to examine interaction effects in $2\times 2$ factorial designs. Experiment 10 used different sorts of residual plots for hierarchical linear models as the visual test statistics.

In experiments 4, 5 and 6, different visual test statistics were used on the same data, providing the ability to examine the effect of these choices. Results are published in \cite{heike:2012} and \cite{XXX}(The violin plot study needs to be referenced).

@@

The visual test statistic is chosen for a particular task. If it is well-designed then departures from the null hypothesis will be visible in the plot, especially if the signal is strong. An important factor that helps an observer to identify the actual data plot in a lineup is the strength of the signal in the data. Simulated data was generated for all of the experiments, except 9, enabling the study of signal strength and actual data plot detection. Primarily the results were as expected, that as the signal strength increased observers more frequently picked the actual data plot as different from the null plots.

%On the other hand, if the null hypothesis is not true, visual test statistic should not display any distinguishable pattern. In fact, some of the null plots may appear to be the most different plot in a lineup influencing the observer to chose a plot different than actual data plot. This is an elegant feature of the lineup. It force the observer to chose a wrong plot when the null hypothesis should not be rejected.

@@

\subsection{Question Design} The researcher knows about the underlying hypothesis but the observer does not necessarily know the underlying details of the lineup. So, the researcher needs to ask a question to the observer to answer while evaluating the lineup. This question should provide the observer a little clue so that the answer reflects the hypothesized patterns in the actual plot. For example Table \ref{tbl:visual_stat} shows the questions asked for the simulation experiments considered for this study. Notice that for case 1 if the observer can identify the actual plot in the lineup that should indicate that the plot chosen has the most vertical difference between groups A and B which is exactly what the researchers intend to examine. Similarly for case 2, a correct evaluation would indicate that the slope is different than the slope that may show up just from randomness.

@@

\subsection{Demographics}  Some of the demographics of the observer such as age, gender, education level and geographical location may have effect on how an observer would examine the lineup. This may produce some variability in the performance of the observer. For example, a high school student may not respond same as a well educated person while evaluating a lineup. There may be variations in different age groups as well. Investigation on subjects with a well variety of age groups is necessary to study this variability.

@@

\subsection{Learning Trend} When an observer evaluates a lineup he or she may learn something from the experiences of the earlier observations. If the same observer is shown another lineup the learning from previous evaluation may help. The more evaluations an observer makes the more skillful the observer may become. This learning trend may or may not be significant overall but it may influence the performance of the observer in some way.

@@

\subsection{Location of Actual Data Plot in the Lineup} The actual data plot is placed in a random spot in a lineup. While evaluating the lineup some people may start looking from some specific part of the lineup. With the help of eye-tracking equipments \citet{zhao:2012} tracked the observers' eyes to see how they went through the plots in a lineup to come to their answers. The results suggest that people have particular methods of reading lineups. Some people read lineups from left to right direction while some read from upward to downward. Some people start looking from the center of the lineup while others start from the top left corner. In the earlier phase of the exercise, the observer tend to scan the plots and start comparing plots to make a final decision. Beside right-left or up-down directions observers show some diagonal movement too.

@@

\subsection{Individual Skill or Ability} Each person is different from others in some way. For example, in a controlled experimental study \citet{zhao:2012}  noticed that some people spent a lot of time to decide no matter whether the lineup is difficult or easy while some simply glanced at lineups to make a decision. This influences the response of the observer. Also subject specific variation in the power of visual test is observed in \cite{majumder:2013}.

@@

\begin{figure}[htbp] 
   \centering
%   \includegraphics[width=6.5in]{learning_trend.pdf} 
    \includegraphics[width=3.0in]{practical_impact_graduate.pdf} 
    \includegraphics[width=3.0in]{practical_impact_degree.pdf} 
   \caption{Proportion of correct responses due to graduate degree as compared to high school degree for an 18-25 year old female in the United States. Even though graduate degree is statistically significant, the largest difference in proportion correct is  0.045 which is very negligible. The difference diminishes as we move away one or two standard deviations ($\sigma_{\ell} =2.293$) of lineup variability.}
   \label{fig:practical_impact_graduate}
\end{figure}
